---
title: "REACH and RStudio: PS and Survival Analysis"
subtitle: "Study designs, outcome models, and running them with Strategus"
author: "Erik Westlund"
date: "2026-10-07"
footer: "BIDS Bootcamp · REACH and RStudio"
execute:
  eval: true
fig-width: 10
fig-height: 4.6
---

<!--
Day 1 is concepts plus live walkthroughs. No student hands-on.
Audience: little background in stats, research design, or the tools.
This is their first explanation of what these models do.

Teaching pattern, repeated for every model:
  1. Simulate clean data where we know the true effect.
  2. Fit the model and read the answer off the output.
  3. Add confounders (age, sex). The unadjusted model gets it wrong.
  4. Adjust for the confounders. We get the right answer back.

Unlike the other decks, this one runs its code at render time so the
simulation output appears on the slides. Requires ggplot2, survival, MatchIt.
-->

```{r}
#| label: setup
#| include: false
library(ggplot2)
library(survival)
library(MatchIt)

theme_set(theme_minimal(base_size = 18))

# Format a number for slide text, with a typographic minus sign
fmt <- function(x, digits = 2) sub("^-", "\u2212", formatC(x, digits = digits, format = "f"))
```

# Welcome

## Today's Plan

1. Study design and causal inference
2. The four regression models used most often in observational research
3. Confounding, and how adjustment removes it
4. A preview of propensity scores and CohortMethod

. . .

On Monday we described cohorts with tables and figures. Today and Friday we move from description to estimation: measuring the effect of a treatment on an outcome.

# Before We Model

## Where This Runs: REACH

- REACH is our deidentified data enclave.
- Analyses on real patient data run in RStudio inside REACH.
- Today's examples use simulated data, so everything shown here runs on a laptop.

## Design Comes First

Aki's lecture covered causal inference. Today we use a few of those ideas.

Before choosing a model, a study needs answers to three questions:

- What is the research question?
- Which groups are being compared?
- What is the outcome, and over what period is it measured?

. . .

A statistical model cannot correct a flawed design. The design determines what the model is able to estimate.

## Why Randomized Trials Work

- In a randomized controlled trial, chance alone decides who receives the treatment.
- As a result, the treated and untreated groups are expected to be the same at baseline in every respect, measured or not.
- Any systematic difference in outcomes between the groups can therefore be attributed to the treatment.

## Observational Data

In observational data, treatment is not assigned at random. Doctors and patients choose treatments for reasons, and those reasons can also affect the outcome.

A **confounder** is a variable that affects both which treatment a person receives and the outcome.

## Confounding by Age and Sex

In today's simulations, older patients and men are more likely to be treated. They are also more likely to have worse outcomes.

```{r}
#| label: dag
#| echo: false
#| fig-height: 3.6
nodes <- data.frame(
  name = c("Age", "Sex", "Treatment", "Outcome"),
  x = c(0.7, 1.3, 0, 2),
  y = c(1, 1, 0, 0)
)
edges <- data.frame(
  from = c("Age", "Age", "Sex", "Sex", "Treatment"),
  to = c("Treatment", "Outcome", "Treatment", "Outcome", "Outcome")
)
edges <- cbind(
  setNames(nodes[match(edges$from, nodes$name), c("x", "y")], c("x", "y")),
  setNames(nodes[match(edges$to, nodes$name), c("x", "y")], c("xend", "yend"))
)
# Shorten each arrow so it stops at the edge of the label
shrink <- 0.22
dx <- edges$xend - edges$x
dy <- edges$yend - edges$y
len <- sqrt(dx^2 + dy^2)
edges <- transform(
  edges,
  x = x + shrink * dx / len, y = y + shrink * dy / len,
  xend = xend - shrink * dx / len, yend = yend - shrink * dy / len
)

ggplot() +
  geom_segment(data = edges, aes(x, y, xend = xend, yend = yend),
               arrow = arrow(length = unit(0.25, "cm"), type = "closed"),
               color = "#7E7E7C", linewidth = 0.8) +
  geom_label(data = nodes, aes(x, y, label = name), size = 7,
             color = "#002D72", linewidth = 0.6,
             label.padding = unit(0.4, "lines")) +
  coord_cartesian(xlim = c(-0.3, 2.3), ylim = c(-0.2, 1.2)) +
  theme_void()
```

A comparison of treated and untreated patients that ignores age and sex mixes the effect of the treatment with the effects of age and sex.

## Why We Simulate

- In a simulation, we choose the true effect of the treatment before generating the data.
- We can then check whether a method recovers that effect.
- Testing a method on simulated data shows how it behaves before we trust it on real data, where the true effect is unknown.

# The Models

## Four Common Models

The right model depends on the type of outcome.

| Outcome | Example | Model | R function | Treatment effect |
|---|---|---|---|---|
| Continuous | Blood pressure | Linear (OLS) | `lm()` | Difference in mean outcome |
| Binary | GI bleed | Logistic | `glm(family = binomial)` | Odds ratio |
| Count or rate | Number of ER visits | Poisson | `glm(family = poisson)` | Rate ratio |
| Time to event | Time until GI bleed | Cox | `coxph()` | Hazard ratio |

## Conditional Effects

Each treatment effect is conditional on the other covariates in the model. It compares treated and untreated patients who have the same values of those covariates.

## Two Simulated Datasets

We build two datasets of 10,000 patients each and reuse them for every model.

- **`trial`**: a randomized trial. A coin flip decides treatment.
- **`obs`**: observational data. Age and sex affect both treatment and the outcome.

. . .

For each model, we add an outcome with a known treatment effect to both datasets. We then check whether the model recovers that effect.

# Linear Regression

## Linear Regression (OLS)

Linear regression models a continuous outcome as a straight-line function of the predictors.

- The **intercept** is the expected outcome when every predictor is zero.
- The **coefficient** on treatment is the expected difference in the outcome between treated and untreated patients who have the same values of the other predictors.

. . .

In the simulation, the outcome is systolic blood pressure, and the treatment lowers it by 5 mmHg.

## Simulating a Randomized Trial

```{r}
#| label: sim-trial
set.seed(21)
n <- 10000

# A coin flip assigns treatment: 1 = treated, 0 = untreated
trial <- data.frame(treated = rbinom(n, size = 1, prob = 0.5))

# Blood pressure averages 140; treatment lowers it by 5; rnorm() adds noise
trial$sbp <- 140 - 5 * trial$treated + rnorm(n, mean = 0, sd = 10)

head(trial, 4)
```

## Comparing the Two Groups

```{r}
#| label: plot-trial-sbp
#| fig-height: 4
ggplot(trial, aes(x = factor(treated), y = sbp)) +
  geom_jitter(width = 0.2, alpha = 0.08, color = "#68ACE5") +
  stat_summary(fun = mean, geom = "point", size = 5, color = "#002D72") +
  scale_x_discrete(labels = c("Untreated", "Treated")) +
  labs(x = NULL, y = "Systolic blood pressure (mmHg)")
```

The dark points mark each group's mean.

## Fitting the Model

`lm()` fits a linear regression. `coef()` returns the estimates, and `confint()` returns their 95% confidence intervals.

```{r}
#| label: fit-ols-trial
fit_ols_trial <- lm(sbp ~ treated, data = trial)
round(cbind(estimate = coef(fit_ols_trial), confint(fit_ols_trial)), 2)
```

. . .

The coefficient on `treated` is close to the true effect of −5. With a single binary predictor, it equals the difference between the two group means:

```{r}
#| label: means-trial
aggregate(sbp ~ treated, data = trial, FUN = mean)
```

## Simulating Observational Data

Now treatment depends on age and sex. `plogis()` converts a score into a probability between 0 and 1.

```{r}
#| label: sim-obs
obs <- data.frame(
  age = rnorm(n, mean = 60, sd = 10),
  male = rbinom(n, size = 1, prob = 0.5)
)

# Older patients and men are more likely to be treated
obs$treated <- rbinom(n, size = 1,
                      prob = plogis(-1.5 + (obs$age - 60) / 10 + obs$male))

aggregate(cbind(age, male) ~ treated, data = obs, FUN = mean)
```

Treated patients are older on average, and a larger share of them are men.

## A Confounded Outcome

Blood pressure rises with age and is higher in men. The treatment effect is still −5.

```{r}
#| label: fit-ols-naive
obs$sbp <- 100 + 0.7 * obs$age + 4 * obs$male - 5 * obs$treated +
  rnorm(n, mean = 0, sd = 10)

fit_ols_naive <- lm(sbp ~ treated, data = obs)
round(cbind(estimate = coef(fit_ols_naive), confint(fit_ols_naive)), 2)
```

. . .

The model estimates an effect of `r fmt(coef(fit_ols_naive)[["treated"]], 1)` mmHg. The treatment appears to raise blood pressure, because treated patients are older and more often men.

## Seeing the Confounding

```{r}
#| label: plot-obs-sbp
#| fig-height: 4.4
ggplot(obs, aes(x = age, y = sbp, color = factor(treated))) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.5) +
  scale_color_manual(values = c("0" = "#7E7E7C", "1" = "#002D72"),
                     labels = c("Untreated", "Treated"), name = NULL) +
  labs(x = "Age", y = "Systolic blood pressure (mmHg)")
```

At any given age, treated patients have lower blood pressure. The treated group is older, which hides this difference when age is ignored.

## Adjusting for Age and Sex

Adding age and sex to the model compares treated and untreated patients of the same age and sex.

```{r}
#| label: fit-ols-adj
fit_ols_adj <- lm(sbp ~ treated + age + male, data = obs)
round(cbind(estimate = coef(fit_ols_adj), confint(fit_ols_adj)), 2)
```

. . .

The adjusted estimate, `r fmt(coef(fit_ols_adj)[["treated"]], 1)` mmHg, is close to the true effect of −5.

# Logistic Regression

## Logistic Regression

Logistic regression models a binary outcome, such as whether a patient has a GI bleed.

- The **odds** of an event are the probability it happens divided by the probability it does not. A 20% risk is odds of 0.2 / 0.8 = 0.25.
- The model estimates coefficients on the **log odds** scale.
- Exponentiating a coefficient with `exp()` gives an **odds ratio**. An odds ratio of 2 means the treatment doubles the odds of the outcome.

. . .

In the simulation, the treatment doubles the odds of a GI bleed.

## Logistic: Randomized Trial

`rbinom()` draws a 0 or 1 for each patient, using the probability from `plogis()`.

```{r}
#| label: fit-logit-trial
trial$bleed <- rbinom(n, size = 1, prob = plogis(-2 + log(2) * trial$treated))

fit_logit_trial <- glm(bleed ~ treated, family = binomial, data = trial)
round(exp(cbind(odds_ratio = coef(fit_logit_trial), confint(fit_logit_trial))), 2)
```

. . .

The odds ratio for `treated` is close to the true value of 2. The intercept row is the odds of a bleed among untreated patients.

## Logistic: Observational, Unadjusted

Older patients and men are also more likely to bleed.

```{r}
#| label: fit-logit-naive
obs$bleed <- rbinom(n, size = 1, prob = plogis(
  -2 + log(2) * obs$treated + 0.05 * (obs$age - 60) + 0.5 * obs$male
))

fit_logit_naive <- glm(bleed ~ treated, family = binomial, data = obs)
round(exp(cbind(odds_ratio = coef(fit_logit_naive), confint(fit_logit_naive))), 2)
```

. . .

The unadjusted odds ratio, `r fmt(exp(coef(fit_logit_naive)[["treated"]]), 2)`, overstates the harm of the treatment.

## Logistic: Observational, Adjusted

```{r}
#| label: fit-logit-adj
fit_logit_adj <- glm(bleed ~ treated + age + male, family = binomial, data = obs)
round(exp(cbind(odds_ratio = coef(fit_logit_adj), confint(fit_logit_adj))), 2)
```

. . .

After adjustment, the odds ratio for `treated` is `r fmt(exp(coef(fit_logit_adj)[["treated"]]), 2)`, close to the true value of 2. The `age` row is the odds ratio for each additional year of age.

# Poisson Regression

## Poisson Regression

Poisson regression models counts of events, such as the number of ER visits.

- Patients are followed for different lengths of time. A patient followed for three years has more time to visit the ER than one followed for six months.
- The model accounts for this with an **offset**: the log of each patient's **person-time**.
- The model then estimates a **rate**, the same quantity as Monday's incidence rates.
- Exponentiating a coefficient gives a **rate ratio**.

. . .

In the simulation, the treatment raises the rate of ER visits by 50%, a rate ratio of 1.5.

## Poisson: Randomized Trial

Each patient is followed for between six months and three years. `rpois()` draws a count of visits from that patient's rate and follow-up time.

```{r}
#| label: fit-pois-trial
trial$years <- runif(n, min = 0.5, max = 3)
trial$visits <- rpois(n, lambda = 0.5 * 1.5^trial$treated * trial$years)

fit_pois_trial <- glm(visits ~ treated + offset(log(years)),
                      family = poisson, data = trial)
round(exp(cbind(rate_ratio = coef(fit_pois_trial), confint(fit_pois_trial))), 2)
```

. . .

The intercept row is the rate among untreated patients: about 0.5 visits per person-year.

## Poisson: Observational, Unadjusted

```{r}
#| label: fit-pois-naive
obs$years <- runif(n, min = 0.5, max = 3)
obs$visits <- rpois(n, lambda = 0.5 * 1.5^obs$treated *
  exp(0.03 * (obs$age - 60) + 0.3 * obs$male) * obs$years)

fit_pois_naive <- glm(visits ~ treated + offset(log(years)),
                      family = poisson, data = obs)
round(exp(cbind(rate_ratio = coef(fit_pois_naive), confint(fit_pois_naive))), 2)
```

. . .

The unadjusted rate ratio is `r fmt(exp(coef(fit_pois_naive)[["treated"]]), 2)`. The true value is 1.5.

## Poisson: Observational, Adjusted

```{r}
#| label: fit-pois-adj
fit_pois_adj <- glm(visits ~ treated + age + male + offset(log(years)),
                    family = poisson, data = obs)
round(exp(cbind(rate_ratio = coef(fit_pois_adj), confint(fit_pois_adj))), 2)
```

. . .

After adjustment, the rate ratio is `r fmt(exp(coef(fit_pois_adj)[["treated"]]), 2)`.

# Survival Analysis

## Survival Analysis

Survival analysis models the time until an event occurs.

- Some patients leave the data before the event can be observed, for example because they change insurance. This is called **censoring**. We know only that they had not had the event by the time they left.
- **Kaplan–Meier curves** show the proportion of each group that is still event-free over time.
- The **Cox model** estimates a **hazard ratio**: the ratio of the event rates in the two groups at any point in time.
- The Cox model assumes that this ratio is constant over time. This is the **proportional hazards** assumption.

. . .

In the simulation, the treatment doubles the hazard of a GI bleed.

## Simulating Time to Event

`rexp()` draws each patient's time to a GI bleed. A second draw sets when each patient leaves the data.

```{r}
#| label: sim-surv-trial
bleed_time <- rexp(n, rate = 0.1 * 2^trial$treated)  # years
censor_time <- runif(n, min = 0, max = 5)

# We observe whichever comes first
trial$time <- pmin(bleed_time, censor_time)
trial$event <- as.integer(bleed_time <= censor_time)  # 1 = bleed, 0 = censored

head(trial[, c("treated", "time", "event")], 4)
```

## Kaplan–Meier Curves

`Surv()` combines the follow-up time and the event indicator. `survfit()` estimates the curves.

```{r}
#| label: km-trial
#| fig-height: 3.8
km <- survfit(Surv(time, event) ~ treated, data = trial)
km_data <- data.frame(time = km$time, surv = km$surv,
                      group = rep(c("Untreated", "Treated"), km$strata))

ggplot(km_data, aes(x = time, y = surv, color = group)) +
  geom_step(linewidth = 1.5) +
  scale_color_manual(values = c(Untreated = "#7E7E7C", Treated = "#002D72"), name = NULL) +
  labs(x = "Years since treatment start", y = "Proportion without a GI bleed")
```

## Cox: Randomized Trial

`coxph()` fits the Cox model. Its output has no intercept.

```{r}
#| label: fit-cox-trial
fit_cox_trial <- coxph(Surv(time, event) ~ treated, data = trial)
round(exp(cbind(hazard_ratio = coef(fit_cox_trial), confint(fit_cox_trial))), 2)
```

. . .

The hazard ratio is close to the true value of 2.

## Cox: Observational, Unadjusted

```{r}
#| label: fit-cox-naive
bleed_time <- rexp(n, rate = 0.1 * 2^obs$treated *
  exp(0.05 * (obs$age - 60) + 0.5 * obs$male))
censor_time <- runif(n, min = 0, max = 5)
obs$time <- pmin(bleed_time, censor_time)
obs$event <- as.integer(bleed_time <= censor_time)

fit_cox_naive <- coxph(Surv(time, event) ~ treated, data = obs)
round(exp(cbind(hazard_ratio = coef(fit_cox_naive), confint(fit_cox_naive))), 2)
```

. . .

The unadjusted hazard ratio is `r fmt(exp(coef(fit_cox_naive)[["treated"]]), 2)`. The true value is 2.

## Cox: Observational, Adjusted

```{r}
#| label: fit-cox-adj
fit_cox_adj <- coxph(Surv(time, event) ~ treated + age + male, data = obs)
round(exp(cbind(hazard_ratio = coef(fit_cox_adj), confint(fit_cox_adj))), 2)
```

. . .

After adjustment, the hazard ratio is `r fmt(exp(coef(fit_cox_adj)[["treated"]]), 2)`.

## Summary

```{r}
#| label: summary-table
#| echo: false
effect <- function(fit, exponentiate) {
  est <- coef(fit)[["treated"]]
  fmt(if (exponentiate) exp(est) else est)
}

knitr::kable(data.frame(
  Model = c("Linear (OLS)", "Logistic", "Poisson", "Cox"),
  `Treatment effect` = c("Difference in means", "Odds ratio", "Rate ratio", "Hazard ratio"),
  Truth = c("−5.00", "2.00", "1.50", "2.00"),
  Trial = c(effect(fit_ols_trial, FALSE), effect(fit_logit_trial, TRUE),
            effect(fit_pois_trial, TRUE), effect(fit_cox_trial, TRUE)),
  `Observational, unadjusted` = c(effect(fit_ols_naive, FALSE), effect(fit_logit_naive, TRUE),
                                  effect(fit_pois_naive, TRUE), effect(fit_cox_naive, TRUE)),
  `Observational, adjusted` = c(effect(fit_ols_adj, FALSE), effect(fit_logit_adj, TRUE),
                                effect(fit_pois_adj, TRUE), effect(fit_cox_adj, TRUE)),
  check.names = FALSE
), align = "llrrrr")
```

In every model, the unadjusted estimate from observational data is biased. Adjusting for the confounders recovers the true effect.

# From Two Confounders to Thousands

## The Problem With Real Data

In the simulations, there were two confounders, and we knew what they were.

In OMOP data, each patient can have thousands of recorded conditions, drugs, procedures, and measurements. Any of them might be a confounder. Adding thousands of covariates to an outcome model is difficult to do well.

## Propensity Scores: A Preview

A **propensity score** is the probability that a patient receives the treatment, given their covariates.

- Patients with similar propensity scores have similar covariates, on average.
- Matching treated and untreated patients on the score produces groups that are comparable at baseline, as in a trial.
- The outcome model can then compare the matched groups directly, without adding the covariates.

## Matching on the Propensity Score

`matchit()` estimates each patient's propensity score from age and sex. It then pairs each treated patient with the untreated patient whose score is closest. The **caliper** sets how close two scores must be. Treated patients with no untreated patient within the caliper are dropped.

```{r}
#| label: ps-match
matched <- matchit(treated ~ age + male, data = obs, caliper = 0.05)
matched_obs <- match.data(matched)

# Treated and untreated patients now have similar age and sex
aggregate(cbind(age, male) ~ treated, data = matched_obs, FUN = mean)
```

## The Outcome Model After Matching

The outcome model contains only the treatment.

```{r}
#| label: ps-outcome
fit_ps <- lm(sbp ~ treated, data = matched_obs)
round(cbind(estimate = coef(fit_ps), confint(fit_ps)), 2)
```

. . .

The estimate is `r fmt(coef(fit_ps)[["treated"]], 1)` mmHg. The true effect is −5, and the unadjusted estimate was `r fmt(coef(fit_ols_naive)[["treated"]], 1)`.

## Friday

- The logic of propensity scores in more detail
- CohortMethod on OMOP data, using the Eunomia GI bleed example
- Diagnostics for checking whether a study can be trusted
- Running the same analysis across a data network with Strategus

## Questions?
