---
title: "Data Characterization & Visualization: Companion Notebook"
author: "Erik Westlund"
date: "2026-10-05"
format:
  html:
    toc: true
    toc-depth: 2
    code-tools: true
    embed-resources: true
    fig-width: 10
    fig-height: 5.6
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

This notebook reproduces every data result and figure in the BIDS bootcamp
talk *Data Characterization & Visualization*, from start to finish. It uses
**Eunomia**, a small synthetic OMOP CDM database that runs entirely on your
own computer, so you can run it without access to any real data.

The notebook has two parts:

1. **Characterization.** Connect to Eunomia, build the cohorts, and run the
   HADES packages (FeatureExtraction and CohortIncidence) plus a few SQL
   queries.
2. **Visualization.** Turn those results into the figures shown in the talk
   with ggplot2.

Each figure is shown below and also saved as a PNG in
`images/characterization/`, which is where the slides read them from.

# Setup

## Packages

The HADES packages come from the OHDSI GitHub organization. Before running this
notebook for the first time, uncomment the install lines and run them once.

```{r}
#| label: packages
# install.packages(c("remotes", "dplyr", "tidyr", "ggplot2", "patchwork", "ragg", "scales"))
# remotes::install_github("OHDSI/Eunomia")
# remotes::install_github("OHDSI/DatabaseConnector")
# remotes::install_github("OHDSI/FeatureExtraction")
# remotes::install_github("OHDSI/CohortIncidence")

library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
```

## Theme and colors

Every figure in the talk uses the Johns Hopkins colors and a theme built on
`theme_minimal()`.

```{r}
#| label: theme
theme_set(theme_minimal())

jhu_colors <- list(
  HopkinsBlue = "#002D72",
  Blue = "#005eb8",
  BlueLight = "#68ACE5",
  PaleBlue = "#A7BCD6",
  Purple = "#470a68",
  LightPurple = "#7c7fab",
  Tan = "#D9C89E",
  Gray1 = "#E5E2E0",
  Gray2 = "#B4B2AD",
  Gray3 = "#7E7E7C",
  Gray4 = "#4A484C",
  Gray5 = "#2C2C33",
  Teal = "#6399AE",
  Green = "#007A53"
)

theme_jhu <- function() {
  theme_minimal(base_family = "Tahoma") +
    theme(
      plot.title = element_text(family = "Georgia", size = 18, hjust = 0.5,
                                face = "bold", margin = margin(b = 10)),
      plot.subtitle = element_text(hjust = 0.5, size = 12, margin = margin(t = 5, b = 15)),
      plot.caption = element_text(size = 10, color = jhu_colors$Gray3, margin = margin(t = 15)),
      axis.text = element_text(size = 9, color = jhu_colors$Gray5),
      axis.title.y = element_text(size = 11, face = "bold", margin = margin(r = 15)),
      axis.title.x = element_text(size = 11, face = "bold", margin = margin(t = 15)),
      axis.line = element_line(color = jhu_colors$Gray1, linewidth = 0.5),
      panel.grid.major.x = element_line(color = jhu_colors$Gray1, linewidth = 0.5),
      panel.grid.minor = element_blank()
    )
}

# Save a figure for the slides, at the size the slides expect
image_dir <- "images/characterization"
dir.create(image_dir, recursive = TRUE, showWarnings = FALSE)

save_figure <- function(plot, name, width = 10, height = 5.6) {
  ggsave(file.path(image_dir, paste0(name, ".png")), plot,
         width = width, height = height, dpi = 200, device = ragg::agg_png)
  plot
}
```

# Part 1: Characterization

## Connect to Eunomia and build the cohorts

`Eunomia::createCohorts()` writes four example cohorts to the `cohort` table.
We use three of them:

| Cohort ID | Cohort |
|---|---|
| 1 | New users of celecoxib |
| 2 | New users of diclofenac |
| 3 | GI bleed |

```{r}
#| label: connect
connection_details <- Eunomia::getEunomiaConnectionDetails()
Eunomia::createCohorts(connection_details)
con <- DatabaseConnector::connect(connection_details)

cohort_names <- c("1" = "Celecoxib", "2" = "Diclofenac", "3" = "GI bleed")
```

## Cohort counts

```{r}
#| label: cohort-counts
cohort_counts <- DatabaseConnector::querySql(
  con,
  "SELECT cohort_definition_id, COUNT(DISTINCT subject_id) AS persons
     FROM main.cohort
    WHERE cohort_definition_id IN (1, 2, 3)
    GROUP BY cohort_definition_id",
  snakeCaseToCamelCase = TRUE
) |>
  mutate(cohort = cohort_names[as.character(cohortDefinitionId)])

cohort_counts
```

## Baseline characteristics (FeatureExtraction)

FeatureExtraction computes the covariates behind a "Table 1": demographics,
plus conditions and drugs recorded in the year before cohort entry. With
`aggregated = TRUE` it returns summary statistics for the whole cohort.

```{r}
#| label: feature-extraction
covariate_settings <- FeatureExtraction::createCovariateSettings(
  useDemographicsGender = TRUE,
  useDemographicsAge = TRUE,
  useDemographicsAgeGroup = TRUE,
  useConditionGroupEraLongTerm = TRUE,
  useDrugGroupEraLongTerm = TRUE,
  longTermStartDays = -365,
  endDays = -1
)

get_covariates <- function(cohort_id) {
  FeatureExtraction::getDbCovariateData(
    connectionDetails = connection_details,
    cdmDatabaseSchema = "main",
    cohortDatabaseSchema = "main",
    cohortTable = "cohort",
    cohortIds = cohort_id,
    covariateSettings = covariate_settings,
    aggregated = TRUE
  )
}
cov_celecoxib <- get_covariates(1)
cov_diclofenac <- get_covariates(2)

smd_all <- FeatureExtraction::computeStandardizedDifference(cov_celecoxib, cov_diclofenac) |>
  as_tibble()
```

The standardized mean difference (SMD) is the difference between the cohorts
on a covariate, in standard deviation units. We keep 20 covariates for the
Table 1. FeatureExtraction computes (comparator − target), so we flip the sign:
a positive SMD means the covariate is more common among celecoxib users.

```{r}
#| label: table1
picked <- tribble(
  ~covariateId, ~covariate,
  8532001,      "Female",
  6003,         "Age 30-34",
  7003,         "Age 35-39",
  8003,         "Age 40-44",
  9003,         "Age 45-49",
  4285898210,   "Polyp of colon",
  30753210,     "Esophagitis",
  81893210,     "Ulcerative colitis",
  4310024210,   "Angiodysplasia of stomach",
  4112343210,   "Acute viral pharyngitis",
  4283893210,   "Sinusitis",
  260139210,    "Acute bronchitis",
  372328210,    "Otitis media",
  81151210,     "Sprain of ankle",
  1125315410,   "Acetaminophen",
  1713332410,   "Amoxicillin",
  1177480410,   "Ibuprofen",
  1115008410,   "Naproxen",
  1112807410,   "Aspirin",
  1322184410,   "Clopidogrel"
)

table1 <- picked |>
  left_join(smd_all, by = "covariateId") |>
  transmute(
    covariate,
    pct_celecoxib = round(100 * mean1, 1),
    pct_diclofenac = round(100 * mean2, 1),
    smd = round(-stdDiff, 3)
  )

table1
```

## Age and sex

FeatureExtraction gives age and sex separately. For the age-by-sex figures we
need the joint distribution, so we query the `cohort` and `person` tables
directly.

```{r}
#| label: age-sex
age_sex_counts <- DatabaseConnector::renderTranslateQuerySql(
  con,
  "SELECT c.cohort_definition_id,
          FLOOR((YEAR(c.cohort_start_date) - p.year_of_birth) / 5) * 5 AS age_group,
          p.gender_concept_id,
          COUNT(*) AS persons
     FROM main.cohort c
     JOIN main.person p ON p.person_id = c.subject_id
    WHERE c.cohort_definition_id IN (1, 2)
    GROUP BY c.cohort_definition_id,
             FLOOR((YEAR(c.cohort_start_date) - p.year_of_birth) / 5) * 5,
             p.gender_concept_id",
  snakeCaseToCamelCase = TRUE
) |>
  mutate(
    cohort = cohort_names[as.character(cohortDefinitionId)],
    sex = if_else(genderConceptId == 8507, "Male", "Female")
  )

# Percent of each cohort in each age group and sex
age_sex <- age_sex_counts |>
  group_by(cohort) |>
  mutate(
    pct = 100 * persons / sum(persons),
    age = paste0(ageGroup, "-", ageGroup + 4),
    pct_signed = if_else(sex == "Female", -pct, pct)  # women's bars point left in the pyramid
  ) |>
  ungroup()

age_sex
```

## Incidence (CohortIncidence)

CohortIncidence computes incidence rates. The design below counts first GI
bleeds in the 365 days after drug start, stratified by age group, sex, and
calendar year.

```{r}
#| label: incidence
ir_design <- CohortIncidence::createIncidenceDesign(
  targetDefs = list(
    CohortIncidence::createCohortRef(id = 1, name = "Celecoxib"),
    CohortIncidence::createCohortRef(id = 2, name = "Diclofenac")
  ),
  outcomeDefs = list(
    CohortIncidence::createOutcomeDef(id = 1, name = "GI bleed", cohortId = 3, cleanWindow = 9999)
  ),
  tars = list(
    CohortIncidence::createTimeAtRiskDef(id = 1, startWith = "start", endWith = "start", endOffset = 365)
  ),
  analysisList = list(
    CohortIncidence::createIncidenceAnalysis(targets = c(1, 2), outcomes = 1, tars = 1)
  ),
  strataSettings = CohortIncidence::createStrataSettings(
    byAge = TRUE, ageBreaks = c(35, 40, 45), byGender = TRUE, byYear = TRUE
  )
)

ir_results <- CohortIncidence::executeAnalysis(
  connectionDetails = connection_details,
  incidenceDesign = ir_design,
  buildOptions = CohortIncidence::buildOptions(
    cohortTable = "main.cohort",
    cdmDatabaseSchema = "main",
    sourceName = "Eunomia GiBleed",
    refId = 1
  )
)
```

The results include every combination of strata. For the figures we keep the
rows stratified by age group and sex (not by year), and give the age groups
readable labels in age order.

```{r}
#| label: incidence-tidy
age_labels <- c("<35" = "30-34", "35 - 39" = "35-39", "40 - 44" = "40-44", ">=45" = "45-49")

ir <- ir_results$incidence_summary |>
  rename_with(tolower) |>
  left_join(ir_results$age_group_def |> rename_with(tolower) |> select(age_group_id, age_group_name),
            by = "age_group_id") |>
  mutate(
    cohort = cohort_names[as.character(target_cohort_definition_id)],
    sex = case_when(gender_id == 8507 ~ "Male", gender_id == 8532 ~ "Female"),
    person_years = person_days / 365.25,
    ir_per_1000py = 1000 * outcomes / person_years
  ) |>
  filter(is.na(start_year), !is.na(age_group_name), !is.na(sex)) |>
  mutate(age_group = factor(age_labels[age_group_name], levels = age_labels))

ir |> select(cohort, sex, age_group, outcomes, person_years, ir_per_1000py)
```

## Time to event

Days from drug start to the first GI bleed after it, for people who bled.

```{r}
#| label: time-to-event
tte <- DatabaseConnector::renderTranslateQuerySql(
  con,
  "SELECT t.cohort_definition_id, t.subject_id,
          MIN(DATEDIFF(DAY, t.cohort_start_date, o.cohort_start_date)) AS days_to_event
     FROM main.cohort t
     JOIN main.cohort o
       ON o.subject_id = t.subject_id
      AND o.cohort_definition_id = 3
      AND o.cohort_start_date >= t.cohort_start_date
    WHERE t.cohort_definition_id IN (1, 2)
    GROUP BY t.cohort_definition_id, t.subject_id",
  snakeCaseToCamelCase = TRUE
) |>
  mutate(cohort = cohort_names[as.character(cohortDefinitionId)])

DatabaseConnector::disconnect(con)

tte |> count(cohort)
```

# Part 2: Visualization

## Building a figure in four steps

Each step adds one layer to the same plot.

```{r}
#| label: ggplot-build
save_step <- function(plot, step) {
  save_figure(plot, paste0("ggplot-build-", step), width = 8, height = 5)
}

# 1. Data and aesthetic mapping: axes, but nothing drawn yet
p1 <- ggplot(ir, aes(x = age_group, y = ir_per_1000py))
save_step(p1, 1)

# 2. Add a geometry: one point per row
p2 <- p1 + geom_point()
save_step(p2, 2)

# 3. Color by sex, connect with lines, and one panel per drug
p3 <- p2 +
  aes(color = sex, group = sex) +
  geom_line() +
  facet_wrap(~cohort)
save_step(p3, 3)

# 4. Colors, labels, and a theme
p4 <- p3 +
  scale_color_manual(values = c(Female = jhu_colors$Purple, Male = jhu_colors$BlueLight)) +
  labs(
    title = "GI bleed incidence by age and sex",
    x = "Age at drug start", y = "Per 1,000 person-years", color = NULL
  ) +
  expand_limits(y = 0) +
  theme_jhu() +
  theme(legend.position = "top", strip.text = element_text(size = 12, face = "bold"))
save_step(p4, 4)
```

## Visualization principles

These figures use small made-up numbers, not Eunomia, so each principle is
easy to see.

### Principle 1: Use a common scale

```{r}
#| label: principle-common-scale
shares <- tibble(
  drug = c("Drug A", "Drug B", "Drug C", "Drug D", "Drug E"),
  pct = c(23, 21, 20, 19, 17)
)

pie <- ggplot(shares, aes(x = "", y = pct, fill = drug)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  scale_fill_manual(values = unname(unlist(jhu_colors[c(
    "HopkinsBlue", "Blue", "BlueLight", "Teal", "Tan"
  )]))) +
  labs(title = "Which drug is used most?", fill = NULL) +
  theme_void(base_family = "Tahoma") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16))

bars <- ggplot(shares, aes(x = pct, y = reorder(drug, pct))) +
  geom_col(fill = jhu_colors$HopkinsBlue, width = 0.7) +
  geom_text(aes(label = paste0(pct, "%")), hjust = -0.2, size = 5) +
  scale_x_continuous(limits = c(0, 27), expand = c(0, 0)) +
  labs(title = "Same data, common scale", x = "Share of users (%)", y = NULL) +
  theme_jhu()

save_figure(pie + bars, "principle-pie-vs-bars")
```

### Principle 2: Start bars at zero

```{r}
#| label: principle-honest-axis
rates <- tibble(group = c("Cohort A", "Cohort B"), rate = c(10.4, 11.1))

truncated <- ggplot(rates, aes(x = group, y = rate)) +
  geom_col(fill = jhu_colors$Purple, width = 0.6) +
  coord_cartesian(ylim = c(10, 11.2)) +
  labs(title = "Axis starts at 10: a big gap?", x = NULL,
       y = "Events per 1,000 person-years") +
  theme_jhu()

full <- ggplot(rates, aes(x = group, y = rate)) +
  geom_col(fill = jhu_colors$HopkinsBlue, width = 0.6) +
  scale_y_continuous(limits = c(0, 12), expand = c(0, 0)) +
  labs(title = "Axis starts at 0: about the same", x = NULL,
       y = "Events per 1,000 person-years") +
  theme_jhu()

save_figure(truncated + full, "principle-honest-axis")
```

### Principle 3: Order categories meaningfully

```{r}
#| label: principle-order
conditions <- tibble(
  condition = c("Anemia", "Asthma", "Depression", "Diabetes", "GERD",
                "Hyperlipidemia", "Hypertension", "Osteoarthritis"),
  pct = c(6, 9, 14, 11, 8, 22, 31, 18)
)

alphabetical <- ggplot(conditions, aes(x = pct, y = factor(condition, levels = rev(sort(condition))))) +
  geom_col(fill = jhu_colors$Gray3, width = 0.7) +
  labs(title = "Alphabetical order", x = "Prevalence (%)", y = NULL) +
  theme_jhu()

sorted <- ggplot(conditions, aes(x = pct, y = reorder(condition, pct))) +
  geom_col(fill = jhu_colors$HopkinsBlue, width = 0.7) +
  labs(title = "Sorted by value", x = "Prevalence (%)", y = NULL) +
  theme_jhu()

save_figure(alphabetical + sorted, "principle-order")
```

### Principle 4: Label directly

```{r}
#| label: principle-direct-labels
drug_colors_4 <- c(
  "Drug A" = jhu_colors$HopkinsBlue, "Drug B" = jhu_colors$BlueLight,
  "Drug C" = jhu_colors$Purple, "Drug D" = jhu_colors$Green
)

uptake <- expand_grid(drug = names(drug_colors_4), year = 2016:2024) |>
  mutate(users = case_when(
    drug == "Drug A" ~ 900 - 40 * (year - 2016),
    drug == "Drug B" ~ 300 + 70 * (year - 2016),
    drug == "Drug C" ~ 600 + 10 * (year - 2016),
    drug == "Drug D" ~ 150 + 25 * (year - 2016)
  ))

with_legend <- ggplot(uptake, aes(x = year, y = users, color = drug)) +
  geom_line(linewidth = 1.1) +
  scale_color_manual(values = drug_colors_4) +
  expand_limits(y = 0) +
  labs(title = "Legend", x = NULL, y = "New users", color = NULL) +
  theme_jhu() +
  theme(legend.position = "right")

with_labels <- ggplot(uptake, aes(x = year, y = users, color = drug)) +
  geom_line(linewidth = 1.1) +
  geom_text(data = filter(uptake, year == max(year)),
            aes(label = drug), hjust = -0.15, size = 4.5, fontface = "bold") +
  scale_color_manual(values = drug_colors_4, guide = "none") +
  scale_x_continuous(limits = c(2016, 2026), breaks = seq(2016, 2024, 2)) +
  expand_limits(y = 0) +
  labs(title = "Labels on the lines", x = NULL, y = "New users") +
  theme_jhu()

save_figure(with_legend + with_labels, "principle-direct-labels")
```

### Principle 5: Use small multiples

```{r}
#| label: principle-small-multiples
set.seed(20261005)
databases <- paste("Database", LETTERS[1:6])
monthly <- expand_grid(database = databases, month = 1:36) |>
  mutate(
    base = c(4, 6, 5, 3, 7, 5)[match(database, databases)],
    trend = c(0.02, -0.03, 0, 0.05, -0.01, 0)[match(database, databases)],
    jump = if_else(database == "Database E" & month >= 18, -3, 0),  # a coding change
    rate = base + trend * month + jump + rnorm(n(), 0, 0.35)
  )

db_colors <- setNames(
  unname(unlist(jhu_colors[c("HopkinsBlue", "BlueLight", "Purple", "Green", "Tan", "Gray3")])),
  databases
)

tangled <- ggplot(monthly, aes(x = month, y = rate, color = database)) +
  geom_line(linewidth = 0.8) +
  scale_color_manual(values = db_colors) +
  expand_limits(y = 0) +
  labs(title = "One panel", x = "Month", y = "Per 1,000 person-years", color = NULL) +
  theme_jhu() +
  theme(legend.position = "bottom", legend.text = element_text(size = 8))

multiples <- ggplot(monthly, aes(x = month, y = rate)) +
  geom_line(color = jhu_colors$HopkinsBlue, linewidth = 0.8) +
  facet_wrap(~database, ncol = 3) +
  expand_limits(y = 0) +
  labs(title = "One panel per database", x = "Month", y = "Per 1,000 person-years") +
  theme_jhu() +
  theme(strip.text = element_text(size = 10, face = "bold"))

save_figure(tangled + multiples, "principle-small-multiples")
```

### Principle 6: Use color to highlight

```{r}
#| label: principle-highlight
sites <- tibble(
  database = c("REACH", paste("Database", LETTERS[1:7])),
  persons = c(4200, 8100, 6500, 5900, 3800, 3100, 2400, 1700)
)

rainbow <- ggplot(sites, aes(x = persons, y = reorder(database, persons), fill = database)) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = unname(unlist(jhu_colors[c(
    "HopkinsBlue", "Blue", "BlueLight", "Purple", "LightPurple", "Teal", "Green", "Tan"
  )])), guide = "none") +
  labs(title = "A color for every bar", x = "Persons in cohort", y = NULL) +
  theme_jhu()

highlight <- ggplot(sites, aes(x = persons, y = reorder(database, persons),
                               fill = database == "REACH")) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = c(`TRUE` = jhu_colors$HopkinsBlue, `FALSE` = jhu_colors$Gray1),
                    guide = "none") +
  labs(title = "Color only for REACH", x = "Persons in cohort", y = NULL) +
  theme_jhu()

save_figure(rainbow + highlight, "principle-highlight")
```

### Principle 7: State the finding in the title

```{r}
#| label: principle-title
bleeds <- tibble(drug = c("Drug A", "Drug B", "Drug C"), rate = c(4.1, 8.3, 4.6))

bleed_bars <- function(title) {
  ggplot(bleeds, aes(x = drug, y = rate)) +
    geom_col(fill = if_else(bleeds$drug == "Drug B", jhu_colors$Purple, jhu_colors$Gray2),
             width = 0.6) +
    scale_y_continuous(limits = c(0, 10), expand = c(0, 0)) +
    labs(title = title, x = NULL, y = "GI bleeds per 1,000 person-years") +
    theme_jhu() +
    theme(plot.title = element_text(size = 14))
}

save_figure(bleed_bars("GI bleed rate by drug") + bleed_bars("Drug B users bleed about twice as often"),
            "principle-title")
```

## Example: cohort sizes

A workhorse first draft with default settings, then a show pony: sorted,
horizontal, labeled, and themed.

```{r}
#| label: counts
p_counts_plain <- ggplot(cohort_counts, aes(x = cohort, y = persons)) +
  geom_col() +
  theme_gray()
save_figure(p_counts_plain, "counts-plain")

p_counts_polished <- ggplot(cohort_counts, aes(x = persons, y = reorder(cohort, persons))) +
  geom_col(fill = jhu_colors$HopkinsBlue, width = 0.6) +
  geom_text(aes(label = scales::comma(persons)), hjust = -0.2, size = 5) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12)), labels = scales::comma) +
  labs(
    title = "Celecoxib users outnumber diclofenac users about 2 to 1",
    subtitle = "People in each cohort, Eunomia GiBleed (synthetic data)",
    x = "People", y = NULL
  ) +
  theme_jhu() +
  theme(axis.text.y = element_text(size = 14), panel.grid.major.y = element_blank())
save_figure(p_counts_polished, "counts-polished")
```

## Example: incidence by age and sex

Three versions: default settings, a workhorse, and a show pony built on the
workhorse.

```{r}
#| label: incidence-defaults
# Defaults only: text age groups sort alphabetically, and four colors to decode
p_ir_plain <- ggplot(ir, aes(x = age_group_name, y = ir_per_1000py, fill = paste(cohort, sex))) +
  geom_col(position = "dodge")
save_figure(p_ir_plain, "incidence-plain")
```

```{r}
#| label: incidence-workhorse
# Workhorse: ages in order, one panel per sex, y axis from zero
p_ir_workhorse <- ggplot(ir, aes(x = age_group, y = ir_per_1000py, color = cohort)) +
  geom_line(aes(group = cohort)) +
  geom_point() +
  facet_wrap(~sex) +
  expand_limits(y = 0)
save_figure(p_ir_workhorse, "incidence-workhorse")
```

```{r}
#| label: incidence-show-pony
# Show pony: the workhorse plus direct labels, colors, titles, and a theme
drug_colors <- c(Celecoxib = jhu_colors$HopkinsBlue, Diclofenac = jhu_colors$Green)

p_ir_polished <- p_ir_workhorse +
  # Label each line at its last point (age 45-49)
  geom_text(data = \(d) filter(d, age_group == "45-49"), aes(label = cohort),
            hjust = -0.25, fontface = "bold", size = 4.5) +
  scale_color_manual(values = drug_colors, guide = "none") +
  # Extra room on the right for the labels
  scale_x_discrete(expand = expansion(add = c(0.4, 1.3))) +
  labs(
    title = "Celecoxib users bled more often in most age-sex groups",
    subtitle = "GI bleeds per 1,000 person-years in the year after drug start (age 45-49: under 10 bleeds per group)",
    x = "Age at drug start", y = "GI bleeds per 1,000 person-years"
  ) +
  theme_jhu() +
  theme(strip.text = element_text(size = 13, face = "bold"))
save_figure(p_ir_polished, "incidence-polished")
```

## Example: Table 1 as a figure

```{r}
#| label: table1-figure
p_smd <- ggplot(table1, aes(x = smd, y = reorder(covariate, smd))) +
  annotate("rect", xmin = -0.1, xmax = 0.1, ymin = -Inf, ymax = Inf,
           fill = jhu_colors$Gray1, alpha = 0.6) +
  geom_vline(xintercept = 0, color = jhu_colors$Gray3) +
  geom_point(aes(color = abs(smd) > 0.1), size = 3.5) +
  scale_color_manual(values = c(`FALSE` = jhu_colors$Gray3, `TRUE` = jhu_colors$HopkinsBlue),
                     guide = "none") +
  labs(
    title = "Celecoxib users had more prior GI disease than diclofenac users",
    subtitle = "Standardized mean difference; shaded band = |SMD| < 0.1 (\"balanced\")",
    x = "SMD (positive = more common in celecoxib)", y = NULL
  ) +
  theme_jhu()
save_figure(p_smd, "table1-smd-plot")
```

## Example: four views of age and sex

The same `age_sex` table, plotted four ways. Only the geoms and facets change.

```{r}
#| label: age-sex-setup
sex_colors <- c(Female = jhu_colors$Purple, Male = jhu_colors$BlueLight)
age_sex_title <- "Age and sex distribution by cohort"
age_sex_subtitle <- "Percent of each cohort by age group and sex at drug start"

theme_age_sex <- function() {
  theme_jhu() +
    theme(legend.position = "top", strip.text = element_text(size = 13, face = "bold"))
}
```

```{r}
#| label: age-sex-stacked
# 1. Stacked bars
p_stacked <- ggplot(age_sex, aes(x = age, y = pct, fill = sex)) +
  geom_col(width = 0.7) +
  facet_wrap(~cohort) +
  scale_fill_manual(values = sex_colors) +
  scale_y_continuous(labels = \(x) paste0(x, "%")) +
  labs(title = age_sex_title, subtitle = age_sex_subtitle,
       x = "Age (years)", y = "Percent of cohort", fill = NULL) +
  theme_age_sex()
save_figure(p_stacked, "age-sex-stacked")
```

```{r}
#| label: age-sex-pyramid
# 2. Pyramid: women's values are negative, so their bars point left
p_pyramid <- ggplot(age_sex, aes(x = pct_signed, y = age, fill = sex)) +
  geom_col(width = 0.8) +
  geom_text(aes(label = sprintf("%.0f%%", pct),
                hjust = if_else(sex == "Female", 1.15, -0.15)), size = 3.8) +
  facet_wrap(~cohort) +
  scale_x_continuous(labels = \(x) paste0(abs(x), "%"), limits = c(-30, 30)) +
  scale_fill_manual(values = sex_colors) +
  labs(title = age_sex_title, subtitle = age_sex_subtitle,
       x = "Percent of cohort", y = "Age (years)", fill = NULL) +
  theme_age_sex()
save_figure(p_pyramid, "age-sex-pyramid")
```

```{r}
#| label: age-sex-faceted
# 3. Facets: one panel per cohort and sex
p_faceted <- ggplot(age_sex, aes(x = pct, y = age, fill = sex)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%.0f%%", pct)), hjust = -0.15, size = 3.8) +
  facet_grid(sex ~ cohort) +
  scale_x_continuous(labels = \(x) paste0(x, "%"), limits = c(0, 30)) +
  scale_fill_manual(values = sex_colors, guide = "none") +
  labs(title = age_sex_title, subtitle = age_sex_subtitle,
       x = "Percent of cohort", y = "Age (years)") +
  theme_age_sex()
save_figure(p_faceted, "age-sex-faceted")
```

```{r}
#| label: age-sex-dotplot
# 4. Connected dot plot: women and men on one scale in each row
p_dotplot <- ggplot(age_sex, aes(x = pct, y = age)) +
  geom_line(aes(group = age), color = jhu_colors$Gray2, linewidth = 1.2) +
  geom_point(aes(color = sex), size = 4) +
  facet_wrap(~cohort) +
  scale_x_continuous(labels = \(x) paste0(x, "%"), limits = c(0, NA)) +
  scale_color_manual(values = sex_colors) +
  labs(title = age_sex_title, subtitle = age_sex_subtitle,
       x = "Percent of cohort", y = "Age (years)", color = NULL) +
  theme_age_sex()
save_figure(p_dotplot, "age-sex-dotplot")
```

## Example: four views of time to GI bleed

```{r}
#| label: tte-setup
# Bleeds per week, by the day each week starts
tte_weekly <- tte |>
  mutate(week_start = 7 * floor(daysToEvent / 7)) |>
  count(cohort, week_start)

# Cumulative percent of each cohort's bleeds by day
tte_cum <- tte |>
  arrange(cohort, daysToEvent) |>
  group_by(cohort) |>
  mutate(pct = 100 * row_number() / n()) |>
  ungroup()

n_users <- setNames(cohort_counts$persons, cohort_counts$cohort)
n_bleeds <- table(tte$cohort)
tte_title <- "Days from drug start to first GI bleed"
tte_subtitle <- sprintf(
  "People with a GI bleed: %d of %s celecoxib users and %d of %s diclofenac users",
  n_bleeds[["Celecoxib"]], scales::comma(n_users[["Celecoxib"]]),
  n_bleeds[["Diclofenac"]], scales::comma(n_users[["Diclofenac"]])
)
day_axis <- scale_x_continuous(breaks = seq(0, 91, by = 14), limits = c(0, 91))
```

```{r}
#| label: tte-histogram
# 1. Histogram: counts per week, one panel per drug, common y axis
p_hist <- ggplot(tte, aes(x = daysToEvent, fill = cohort)) +
  geom_histogram(binwidth = 7, boundary = 0, closed = "left", color = "white") +
  facet_wrap(~cohort, ncol = 1) +
  scale_fill_manual(values = drug_colors, guide = "none") +
  day_axis +
  labs(title = tte_title, subtitle = tte_subtitle,
       x = "Days since drug start", y = "People (per week)") +
  theme_jhu() +
  theme(strip.text = element_text(size = 13, face = "bold", hjust = 0))
save_figure(p_hist, "time-to-event-histogram")
```

```{r}
#| label: tte-lines
# 2. Weekly counts as connected lines on one panel
p_lines <- ggplot(tte_weekly, aes(x = week_start, y = n, color = cohort)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  geom_text(data = \(d) slice_max(d, week_start, by = cohort),
            aes(label = cohort), hjust = -0.15, fontface = "bold", size = 4.5) +
  scale_color_manual(values = drug_colors, guide = "none") +
  scale_x_continuous(breaks = seq(0, 91, by = 14), limits = c(0, 100)) +
  expand_limits(y = 0) +
  labs(title = tte_title, subtitle = tte_subtitle,
       x = "Days since drug start (week starting)", y = "People (per week)") +
  theme_jhu()
save_figure(p_lines, "time-to-event-lines")
```

```{r}
#| label: tte-facets
# 3. Weekly counts, one panel per drug, common y axis
p_tte_facets <- ggplot(tte_weekly, aes(x = week_start, y = n, color = cohort)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  facet_wrap(~cohort, ncol = 1) +
  scale_color_manual(values = drug_colors, guide = "none") +
  day_axis +
  expand_limits(y = 0) +
  labs(title = tte_title, subtitle = tte_subtitle,
       x = "Days since drug start (week starting)", y = "People (per week)") +
  theme_jhu() +
  theme(strip.text = element_text(size = 13, face = "bold", hjust = 0))
save_figure(p_tte_facets, "time-to-event-facets")
```

```{r}
#| label: tte-cumulative
# 4. Cumulative percent of each cohort's bleeds
p_cum <- ggplot(tte_cum, aes(x = daysToEvent, y = pct, color = cohort)) +
  geom_step(linewidth = 1) +
  geom_text(data = \(d) slice_max(d, daysToEvent, by = cohort, with_ties = FALSE),
            aes(label = cohort), hjust = -0.1, vjust = c(1.5, -0.5),
            fontface = "bold", size = 4.5) +
  scale_color_manual(values = drug_colors, guide = "none") +
  scale_x_continuous(breaks = seq(0, 91, by = 14), limits = c(0, 105)) +
  scale_y_continuous(labels = \(x) paste0(x, "%"), limits = c(0, 100)) +
  labs(title = tte_title, subtitle = tte_subtitle,
       x = "Days since drug start", y = "Cumulative percent of bleeds") +
  theme_jhu()
save_figure(p_cum, "time-to-event-cumulative")
```

# Session info

```{r}
#| label: session-info
sessionInfo()
```
